### 2018-07-02 paper/6-engineering.tex
Work on S6

While we have given simple implementations assuming a single thread of execution,
all the build systems we address can actually build independent keys in parallel.
While it complicates the model, the complications can be restricted exclusively
to the dependency strategy:

\begin{enumerate}
\item The \hs{topological} function can build the full dependency graph, and
whenever all dependencies of a task are complete, the task itself can be started.

\item The \hs{reordering} function can be made parallel in a few ways, but the
most direct is to have $n$ threads reading entries from the list of keys. As
before, if a key requires something not yet built, it is added to the end -- the
difference is that sometimes things will be moved to the back of the queue not
because they are out of order but because of races with earlier nodes that had
not yet finished. As a consequence, over successive runs, potentially racey
dependencies will be separated, giving better parallelism over time.

\item The \hs{recursive} function can be made parallel by starting static
dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
resolved, using the strategy described by~\citet{marlow2014haxl}.
\end{enumerate}

The actual implementation of the parallel strategies is not overly onerous,
but neither is it beautiful or informative.

@@

In our model we define \hs{Task} as a function -- when given the same inputs it
will always produce the same outputs. Alas, the real-world is not so obliging.
Some examples of impure tasks include:

@@

Most of the implementations in \S\ref{sec-implementations} can deal with non-determinism,
apart from \Bazel, which requires deterministic execution, and in turn can
optimise the number of roundtrips required to the server.

One tempting way of modelling non-determinism is to enrich \hs{Task} from
\hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
respectively. More concretely, the following task description corresponds to
a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh3 :: Task MonadPlus String Integer
sprsh3 fetch "B1" = Just $ (+) <$> fetch "A1" <*> pure 1 `mplus` pure 2
sprsh3 _     _    = Nothing
\end{minted}
\vspace{1mm}

\noindent
Handling such tasks is possible in our framework, but requires an adjustment of
the correctness definition (\S\ref{sec-build-correctness}): instead of requiring
that the result of recomputing the \hs{task} matches the value stored in the \hs{result}:
\[
\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}~\hs{k}\]
\noindent
we now require that \hs{result} contains \emph{one possible result of recomputing}
the \hs{task}:
\[
\hs{Just}~\hs{(}\hs{getValue}~\hs{k}~\hs{result)}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}~\hs{k}
\]
where
\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{k}~\hs{->}~\hs{Maybe}~\hs{[@@v]}
returns the list of all possible results of the \hs{task} instead of just one
value (`ND' stands for `non-deterministic').
% (a possible implementation chooses
% \hs{f}~\hs{=}~\hs{ListT}~\hs{Identity}~\hs{[@@v]}).

@@

\item[Communication] When traces or contents are stored on a central server
communication can become a bottleneck, so it is important to send only the
minimum amount of information, optimising with respect to build-system specific
invariants~--~see~\S\ref{sec-smart-traces} for some possible optimisations.

@@

To legitimise shallow builds, we need to relax the correctness
Definition~\ref{def-correct} as follows. Let the \hs{shallow} store correspond
to the result of a shallow build. Then \hs{shallow} is correct, if \emph{there
exists} \hs{result} which satisfies all requirements of
Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with
\hs{result} on the input keys \hs{k}~$\in$~$I$:
\[
\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}
\]
and on the target \hs{key}:
\[
\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result}.
\]

@@

\noindent
In \Excel the user can alter the value produced by \cmd{B1} by either editing
the inputs of \cmd{A1} or \cmd{A2}, \emph{or} editing the formula in
\cmd{B1}~--~e.g. to \cmd{A1 - A2}. This pattern can be captured by describing
the rule producing \cmd{B1} as also depending on the value \cmd{B1-formula}.
The implementation can be given very directly in a \hs{Task}~\hs{Monad} as:

\begin{minted}{haskell}
task fetch "B1" = do
    formula <- fetch "B1-formula"
    evalFormula fetch formula
\end{minted}

\noindent
Namely, first look up the formula, then interpret it. It is not possible to
change dependencies based on the formula in a \hs{Task}~\hs{Applicative}, as
would be required, so instead the formula can be captured as a dependency
(but its value not used) and \emph{also} baked directly into the \hs{task}
function.

>>>>

While we have given simple implementations assuming a single thread of execution,
all the build systems we address can actually build independent keys in parallel.
While it complicates the model, the complications can be restricted exclusively
to the scheduler:

\begin{enumerate}
\item The \hs{topological} scheduler can build the full dependency graph, and
whenever all dependencies of a task are complete, the task itself can be
started.

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading entries from the list of keys. As
before, if a key requires something not yet built, it is added to the end -- the
difference is that sometimes things will be moved to the back of the queue not
because they are out of order but because of races with earlier nodes that had
not yet finished. As a consequence, over successive runs, potentially racey
dependencies will be separated, giving better parallelism over time.

\item The \hs{suspending} function can be made parallel by starting static
dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
resolved, using the strategy described by~\citet{marlow2014haxl}.
\end{enumerate}

The actual implementation of the parallel schedulers is not overly onerous,
but neither is it beautiful or informative.

@@

In our model we define a \hs{Task} as a function -- when given the same inputs
it will always produce the same output. Alas, the real-world is not so obliging.
Some examples of impure tasks include:

@@

Most of the implementations in \S\ref{sec-implementations} can deal with
non-determinism, apart from \Buck, which relies on deterministic tasks, and
in turn can optimise the number of roundtrips required to the server.

One tempting way of modelling non-determinism is to enrich \hs{Tasks} from
\hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
respectively. More concretely, the following task description corresponds to
a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh3 :: Tasks MonadPlus String Integer
sprsh3 "B1"@\,@=@\,@Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure 1 <|> pure 2)
sprsh3 _   @\,@=@\,@Nothing
\end{minted}
\vspace{1mm}

\noindent
Handling such tasks is possible in our framework, but requires an adjustment of
the correctness definition (\S\ref{sec-build-correctness}): instead of requiring
that the result of recomputing the \hs{task} matches the value stored in the
\hs{result}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
\]
\noindent
we now require that \hs{result} contains \emph{one possible result of recomputing}
the \hs{task}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result}
\]
where
\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{[@@v]}
returns the list of all possible results of the \hs{task} instead of just one
value (`ND' stands for `non-deterministic').

@@

\item[Communication] When traces or contents are stored on a central server
communication can become a bottleneck, so it is important to send only the
minimum amount of information, optimising with respect to build-system specific
invariants.

@@

To legitimise shallow builds, we need to relax the correctness
Definition~\ref{def-correct} as follows. Let the \hs{shallow} store correspond
to the result of a shallow build. Then \hs{shallow} is correct, if \emph{there
exists} a \hs{result} which satisfies all requirements of the
Definition~\ref{def-correct}, \emph{such that} \hs{shallow} agrees with the
\hs{result} on all the input keys \hs{k}~$\in$~$I$:
\[
\hs{getValue}~\hs{k}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{k}~\hs{result}
\]
and on the target \hs{key}:
\[
\hs{getValue}~\hs{key}~\hs{shallow}~\hs{==}~\hs{getValue}~\hs{key}~\hs{result}.
\]

@@

\noindent
In \Excel the user can alter the value produced by \cmd{B1} by either editing
the inputs of \cmd{A1} or \cmd{A2}, \emph{or} editing the formula in
\cmd{B1}~--~e.g. to \cmd{A1 - A2}. This pattern can be captured by describing
the rule producing \cmd{B1} as also depending on the value \cmd{B1-formula}.
The implementation can be given very directly in a \hs{Tasks}~\hs{Monad} as:

\begin{minted}{haskell}
sprsh5 "B1" = Just $ Task $ \fetch -> do
    formula <- fetch "B1-formula"
    evalFormula fetch formula
\end{minted}

\noindent
Namely, first look up the formula, then interpret it. It is not possible to
change dependencies based on the formula in a \hs{Tasks}~\hs{Applicative}, as
would be required, so instead the formula can be captured as a dependency
(but its value not used) and \emph{also} baked directly into the \hs{sprsh5}
function.
